### 2022-09-22 AISTATS2023/paper.tex
updated synthetic examples 1

@@

Recently, ML has flourished in critical domains, such as healthcare
and finance. In these areas, we need ML system with the capability to
explain their predictions, apart from predicting with accuracy. For
this reason there is an increased interest in Explainable AI (XAI),
the field that provides interpretations about the behavior of complex
black-box models. XAI literature distinguishes between local and
global explainability
techniques~\citep{Molnar2020interpretable}. Local methods explain a
specific prediction, whereas global methods explain the entire model
behavior. Global methods provide a universal explanation, summarizing
the numerous local ones into a single interpretable outcome, usually a
number or a plot. If a user wants to get a rough overview about which
features are significant (feature importance) or whether a particular
feature has a positive or negative effect on the output (feature
effect), they should opt for a global explainability technique. On
the other hand, aggregating the individual explanations for producing
a concise global one is vulnerable to misinterpretations; Under strong
feature interactions, the global explanation may obfuscate
heterogeneous effects~\citep{Herbinger2022repid} that exist under the
hood; a phenomenon called aggregation bias~\citep{mehrabi2021survey}.

@@

\clearpage
\section{EXPERIMENTS}

The experimental section is divided in simulation and real-world
examples. The simulation examples, where the data-generating
distribution \(p(\mathbf{X})\) and the predictive function
\(f(\cdot)\) are defined by us, are used for comparing all methods
against a solid ground-truth. In particular, we design two end-to-end
simulation examples; At the first one, we focus on comparing NAME
method versus PDP-ICE in terms of quantifying the heterogeneous
effects for a range of different black-box functions. At the second
one, we provide empirical clues that NAME method achieves a better
approximation compared to traditional ALE, due to appropriate
bin-splitting. Afterwards, we also apply the methods on some
real-world datasets. In these cases, it is infeasible to obtain ground
truth information, therefore, we perform evaluation at two
levels. First, we compare visually NAME versus PDP with ICE
plots. Second, we hold as ground-truth the effects computed on the
full training-set and we, then, perform subsampling to evaluate the
robustness of the bin-splitting method.

\paragraph{Metrics.} Describe the metrics used for comparing methods.

\subsection{Synthetic Example 1: Uncertainty Quantification}

This synthetic example aims to showcase the superiority of NAME method
in quantifying the main effect and the uncertainty compared to PDP
with ICE plots. For being able to compare the two methods against a
solid ground-truth, we limit ourserlves a simple evaluation framework based
on following piecewise linear function:

\begin{equation}
  \label{eq:synth-ex-1-function}
  f(\mathbf{x}) = \begin{cases}
                    f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $f_{\mathtt{lin}} < 0.5$ }\\
                    0.5 - f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $0.5 \leq f_{\mathtt{lin}} < 1$}\\
                    \alpha f_{\mathtt{int}} &\text{otherwise}
                  \end{cases}
\end{equation}
%
where \(f_{\mathtt{lin}}(\mathbf{x}) = a_1 x_1 + a_2 x_2\) and
\(f_{int}(\mathbf{x}) = x_1x_3\). As we observe, the linear term
\(f_{\mathtt{lin}}\) includes the two correlated features and the term
\(f_{\mathtt{int}}\) interacts the two non-correlated variables. The
samples that we use in all examples are coming from the following
distribution: \(p(\mathbf{x}) = p(x_3)p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\),
\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0)\) and
\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\). We will test three variations
by changing the parameters of the black-box function and of the
generative distribution. Despite the model's simplicity, as we will
see below, PDP with ICE plot fail completely to capture the
ground-truth effect, due to the correlation between features \(X_1\)
and \(X_2\). Furthermore, besides not the focus of the current
example, NAME method manages to automatically extract the regions with
constant effect. Details for obtaining the ground truth effects and
the experimental set-up are in the Appendix.\footnote{The Appendix
  contains the effects for all the features, some of them are ommited
  here for compact presentation.}

@@

We test the feature effect when no interaction term is apparent,
i.e. \(\alpha=0\), and the weights are equal \(a_1=a_2=1\).  In this
case, ground truth effect for \(f^{\mathtt{GT}}(x_1)\) is \(x_1\) when
\(0 \leq x_1 < 0.25\), \(-x_1\) when \(0.25 \leq x_1 < 0.5\) and zero
otherwise. In each position \(x_1\), there is complete absence of
heterogeneous effects, and therefore, the uncertainty of the global
explanation is zero, i.e. \(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). In
Figure~\ref{fig:ex-synth-1-1} we see that PDP effect is completely
wrong and ICE plots imply heterogeneous effects. In contrast, ALE
captures correctly the average effect and the zero uncertainty.

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_1/dale_feat_0.png}
  \includegraphics[width=.23\textwidth]{example_1/pdp_ice_feat_0.png}
  \caption{Feature effect for \(x_1\); NAME method at the left and PDP
    with ICE at the right.}
  \label{fig:synth-ex-1-case-1}
\end{figure}

\paragraph{No Uncertainty, Different weights.}

As before, there is no interaction term, i.e. \(\alpha=0\), and
uncertainty about the value of \(x_2\), i.e.  \(x_2 = x_1\), but we
give different weight to each feture \(a_1=2, a_2=0.5\). Again, we
observe that ALE quantifies perfectly the ground truth effect and the
zero-uncertainty in contrast to PDP with ICE.

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_2/dale_feat_0.png}
  \includegraphics[width=.23\textwidth]{example_2/pdp_ice_feat_0.png}
  \caption{Figure 1}
  \label{fig:ex-synth-1-1}
\end{figure}

@@

In this case we activate the interaction term, i.e. \(a=1\), and we
set equal weights \(a_1=a_2=1\). In contrast with above, the
interaction term provokes heterogeneous effects for features
\(x_1, x_3\); at any position \(x_1\), the local effects dependend on
the unknown value of \(X_3\) and vice-versa. The effect of \(x_2\)
continues to have zero-uncertainty, since it does not appear in any
interaction term. NAME captures perfectly the effect of all features,
and the uncertainty in the case of \(x_1, x_2\). On the contrary,
PDP-ICE quantify the effect and the uncertainty correctly only in the
case of \(x_3\). For all other features, both the average effect and
the uncertainty implied by ICE plots are wrong.

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_3/dale_feat_0.png}
  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_0.png}\\
  \includegraphics[width=.23\textwidth]{example_3/dale_feat_1.png}
  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_1.png}\\
  \includegraphics[width=.23\textwidth]{example_3/dale_feat_2.png}
  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_2.png}\\
  \caption{Figure 1}
  \label{fig:ex-synth-1-1}
\end{figure}

@@

The examples presented above, do not cover all cases. Therefore, they
are not sufficiect to prove that NAME method quantifies the effect
correctly in all cases. Their main objective is to highlight that PDP
indifference for correlated features provoques erroneous effects, even
in very simple cases. Unfortunately, the methods that are based on
PDP and ICE to quantify heterogeneity are also vulnerable to the same
errors.

The examples above do not cover the case where the interaction term
include correlated features, \(x_1, x_2\) for our example. In this
case, there is controversy about the ground-truth
effect~\citep{Gromping2020MAEP}.

\subsection{Synthetic Example 2: Bin-Splitting}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_1.jpg}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_2.jpg}\\
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_3.jpg}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_4.jpg}
  \caption{Figure 1}
  \label{fig:ex-synth-1-1}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_1.jpg}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_2.jpg}\\
  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_3.jpg}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_4.jpg}
  \caption{Figure 1}
  \label{fig:ex-synth-1-1}
\end{figure}

@@

>>>>

\usepackage{tikz}
\usetikzlibrary{matrix,positioning,arrows.meta,arrows,fit,backgrounds,decorations.pathreplacing}
\tikzset{ mymat/.style={ matrix of math nodes, text height=2.5ex, text
depth=0.75ex, text width=6.00ex, align=center, column
sep=-\pgflinewidth, nodes={minimum height=5.0ex} }, mymats/.style={
mymat, nodes={draw,fill=#1} }, mymat2/.style={ matrix of math nodes,
text height=1.0ex, text depth=0.0ex, minimum width=5ex, % text
width=7.00ex, align=center, column sep=-\pgflinewidth }, }
\usetikzlibrary{shapes.geometric, arrows, backgrounds, scopes}
\usepackage{pgfplots} \pgfplotsset{width=6.75cm, compat=newest}
\usepackage[utf8]{inputenc} \DeclareUnicodeCharacter{2212}{−}
\usepgfplotslibrary{groupplots,dateplot}
\usetikzlibrary{patterns,shapes.arrows}

@@

Recently, ML has flourished in critical domains, such as healthcare
and finance. In these areas, we need ML system with the capability to
explain their predictions, apart from predicting with accuracy. For
this reason there is an increased interest in Explainable AI (XAI),
the field that provides interpretations about the behavior of complex
black-box models. XAI literature distinguishes between local and
global explainability
techniques~\citep{Molnar2020interpretable}. Local methods explain a
specific prediction, whereas global methods explain the entire model
behavior. Global methods provide a universal explanation, summarizing
the numerous local ones into a single interpretable outcome, usually a
number or a plot. If a user wants to get a rough overview about which
features are significant (feature importance) or whether a particular
feature has a positive or negative effect on the output (feature
effect), they should opt for a global explainability technique. On
the other hand, aggregating the individual explanations for producing
a concise global one is vulnerable to misinterpretations; Under strong
feature interactions, the global explanation may obfuscate
heterogeneous effects~\citep{herbinger2022repid} that exist under the
hood; a phenomenon called aggregation bias~\citep{mehrabi2021survey}.

@@

\clearpage
\section{SIMULATION EXAMPLES}
\label{sec:simulation-examples}

The simulation examples, where the data-generating distribution
\(p(\mathbf{X})\) and the predictive function \(f(\cdot)\) are defined
by us, enable the evaluation of competitive approaches against a solid
ground-truth. We follow this common XAI
practice~\citep{aas2021explaining, herbinger2022repid} for providing
secure empirical evidence about the superiority of NAME method.

We split the simulation examples in two groups. The first group,
Section~\ref{sec:simulation-examples-1}, aims at showcasing that NAME
method is more accurate than PDP-ICE in quantifying the average effect
and the level of heterogeneous effects (uncertainty), when input
features are correlated. The second group,
Section~\ref{sec:simulation-examples-2}, illustrates that NAME method
achieves a better approximation (average effect and uncertainty) in
cases of limited samples, due to automatic bin-splitting. In both
groups, we choose to illustrate the most indicative examples; a more
extensive evaluation is provided in the Appendix.

\subsection{Case 1: Uncertainty Quantification}
\label{sec:simulation-examples-1}

In this simulation, we will compare NAME method against PDP-ICE in
quantifying the main effect \(f_\mu\) and the uncertainty
\(f_\sigma^2\). Since there is ambiguity about the ground-truth
first-order effect in cases of correlated
features, e.g.~\citep{apley2020visualizing,Gromping2020MAEP}, we limit
ourselves to the following piecewise linear function,

\begin{equation}
  \label{eq:synth-ex-1-function}
  f(\mathbf{x}) = \begin{cases}
                    f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $f_{\mathtt{lin}} < 0.5$ }\\
                    0.5 - f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $0.5 \leq f_{\mathtt{lin}} < 1$}\\
                    \alpha f_{\mathtt{int}} &\text{otherwise}
                  \end{cases}
\end{equation}
%
where \(f_{\mathtt{lin}}(\mathbf{x}) = a_1 x_1 + a_2 x_2\) and
\(f_{int}(\mathbf{x}) = x_1x_3\). As we observe, the linear term
\(f_{\mathtt{lin}}\) includes the two correlated features and the term
\(f_{\mathtt{int}}\) interacts the two non-correlated variables. The
samples that we use in all examples are coming from the following
distribution: \(p(\mathbf{x}) = p(x_3)p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\), \(x_2 = x_1\) and
\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\). We will test the effect
computed by NAME and PDP-ICE in three cases; (a) no interaction
(\(\alpha=0\)) and equal weights (\(a_1=a_2\)), (b) no interaction
(\(\alpha=0\)) and different weigths (\( a_1 \neq a_2 \)) and (c) with
interaction (\(\alpha \neq 0\)) with equal weights (\(a_1=a_2\)).

In all cases, we firstly compute the ground-truth average effect and
the uncertainty analytically (proofs in the Appendix) and then we
compare it against the approximation provided by each method. The
approximation is computed after generating \(N=300\) samples. As we
will see, despite the model's simplicity, PDP-ICE fail in quantifying
the effect correctly, due to correlation between features \(X_1\) and
\(X_2\). Finally, besides it is not the focus of the current examples,
we will also observe that NAME manages to correctly extract the
regions with constant effect. Details for obtaining the ground truth
effects and the experimental set-up are in the Appendix.

@@

We test the feature effect when no interaction term is apparent,
i.e. \(\alpha=0\), and the weights are equal \(a_1=a_2=1\).  In this
case, the ground truth effect \(f_\mu^{\mathtt{GT}}(x_1)\) is: \(x_1\)
when \(0 \leq x_1 < 0.25\), \(-x_1\) when \(0.25 \leq x_1 < 0.5\) and
zero otherwise. In each position, there is total absence of
heterogeneous effects; therefore, the uncertainty of the global
explanation is zero \(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). In
Figure~\ref{fig:ex-synth-1-1}, we observe that PDP main effect is
wrong and ICE plots imply the existence of heterogeneous effects. In
contrast, ALE captures correctly the average effect and the zero
uncertainty and it also groups perfectly the regions with
constant-effect.

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_1/dale_feat_0.pdf}
  \includegraphics[width=.23\textwidth]{example_1/pdp_ice_feat_0.pdf}
  \caption{No interaction, Equal weights: Feature effect for \(x_1\);
    Ground-truth vs (a) NAME method at the left and (b) PDP-ICE at the
    right.}
  \label{fig:synth-ex-1-case-1}
\end{figure}

\paragraph{No Interaction, Different weights.}

As before, there is no interaction term and, therefore, the
ground-truth is zero, i.e.  \(f^{\mathcal{GT}}_{\sigma^2}(x_1) =
0\). The weigts are different \(a_1=2, a_2=0.5\), therefore, the
ground-truth effect is \(f_\mu^{\mathtt{GT}}(x_1)\) is: \(2x_1\) when
\(0 \leq x_1 < 0.2\), \(-2x_1\) when \(0.2 \leq x_1 < 0.4\) and zero
otherwise. In Figure~\ref{fig:ex-synth-1-2}, we observe that PDP
estimation is completely opposite to the ground-truth effect,
i.e.~negative in \([0, 0.2)\) and positive in \([0.2, 0.4)\), and the
ICE erroneously implies heterogeneous effects.  As before, ALE
quantifies perfectly the ground truth effect and the zero-uncertainty,
extracting correcltly the constant effect regions.

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_2/dale_feat_0.pdf}
  \includegraphics[width=.23\textwidth]{example_2/pdp_ice_feat_0.pdf}
  \caption{No interaction, Different weights: Feature effect for \(x_1\);
    Ground-truth vs (a) NAME method at the left and (b) PDP-ICE at the
    right.}
  \label{fig:ex-synth-1-2}
\end{figure}

@@

In this case we activate the interaction term, i.e. \(a=1\), and we
set equal weights \(a_1=a_2=1\). The interaction term provokes
heterogeneous effects in features \(x_1, x_3\), i.e., at any position
\(x_1\), the local effects dependend on the unknown value of \(X_3\)
and vice-versa. The effect of \(x_2\) continues to have
zero-uncertainty, since it does not appear in any interaction term. As
we observe in Figure~\ref{fig:ex-synth-1-3}, NAME captures perfectly
the effect of all features, and the uncertainty in all cases. As
expected, PDP-ICE quantify the effect and the uncertainty correctly
only in the case of \(x_3\), since \(X_3\) is independent from other
features. For the correlated features, i.e. \(x_1, x_2\), the average
effect computed by PDP and the uncertainty implied by ICE plots are
wrong.

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_3/dale_feat_0.pdf}
  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_0.pdf}\\
  \includegraphics[width=.23\textwidth]{example_3/dale_feat_1.pdf}
  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_1.pdf}\\
  \includegraphics[width=.23\textwidth]{example_3/dale_feat_2.pdf}
  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_2.pdf}\\
  \caption{With interaction, equal weights: Feature effect for all
    features, \(x_1\) to \(x_3\) from top to bottom; Ground-truth vs
    (a) NAME method at the left columns and (b) PDP-ICE at the right
    column.}
  \label{fig:ex-synth-1-3}
\end{figure}

@@

In purpose, the examples above do not cover the case where there is an
interaction term includes correlated features, i.e. \(x_1x_2\). In
this scenario, there is an open debate about the ground-truth
effect~\citep{Gromping2020MAEP}.

\subsection{Examples 2: Bin-Splitting}
\label{sec:simulation-examples-2}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_1.pdf}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_2.pdf}\\
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_3.pdf}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_4.pdf}
  \caption{Figure 1}
  \label{fig:ex-synth-1-1}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_1.pdf}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_2.pdf}\\
  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_3.pdf}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_4.pdf}
  \caption{Figure 1}
  \label{fig:ex-synth-1-1}
\end{figure}

@@

we also
apply the methods on some real-world datasets. In these cases, it is
infeasible to obtain ground truth information, therefore, we perform
evaluation at two levels. First, we compare visually NAME versus PDP
with ICE plots. Second, we hold as ground-truth the effects computed
on the full training-set and we, then, perform subsampling to evaluate
the robustness of the bin-splitting method.
